---
title: "Spatial Downscaling of Alien Species Presences using Machine Learning"
author: "Ioannis N Daliakopoulos*, Stelios Katsanevakis and Aristides Moustakas"
date: '*Correspondance: daliakopoulos@hydromech.gr'
output:
  word_document: default
  pdf_document: default
  html_document: default
fig_caption: yes
---

```{r Options, echo=F}
rm(list=ls()) #clear workspace
knitr::opts_chunk$set(cache=TRUE, fig.width=12, fig.height=8,
                      echo=T, warning=FALSE, message=FALSE)
options(digits=3, width = 600)
```
Load required libraries
```{r Libraries}
library(rgdal)
library(raster)
library(R2BayesX)
library(tidyr)
library(party)
library(randomForest)
library(caret)
library(ROCR)
library(e1071)
library(matrixStats)
library(vita)
library(ggplot2)
library(Hmisc)
library(gridExtra)
```
Initialize plotting function for maps
```{r Plotting functions for maps}
plot_map = function(p, x, title){
  plot(p, main=title)
  image(x, col=rev(heat.colors(255)), add=TRUE)
  plot(p, add=TRUE)
  tick_locations = c(minValue(x), maxValue(x))
  colorlegend(color = heat.colors, swap=TRUE, plot=FALSE, 
              at= tick_locations, pos=c(0.55,0.9), add=TRUE, range = tick_locations)
  axis(1, tck=.02, at=x<-seq(450000, 700000, by=50000 ), 
       labels=c(450, 500, 550, 600, 650, "700 km"))
  axis(2, tck=.02, at=x<-seq(3850000, 3950000, by=50000 ), 
       labels=c("3,850", "3,900", "3,950 km"))
  box()  
}
```
Initialize plotting function for partial plots
```{r Plotting functions for partials}
plot_partial = function(data, range, ylab, xlab){
  alpha_value <- 0.5 
  p<-ggplot(data=data, aes(x=x, y=y)) +
    geom_point(size=1, alpha=alpha_value, colour = "grey30") +
    geom_smooth(colour = "black", method = "lm", formula = y ~ splines::bs(x, 3))+
    geom_line(colour = "red", data=data[range,]) + 
    theme_bw()+
    theme(axis.ticks.length = unit(-0.05 , "cm"))+ 
    theme(axis.text.x = element_text(margin=margin(3,3,3,3,"pt")))+
    theme(axis.text.y = element_text(margin=margin(3,3,3,3,"pt")))+
    ylab(ylab)+
    xlab(xlab)+ 
    ylim(0, 1)
  p$labels$colour<-""
  p
}
```
Initialize plotting function for shaded areas
```{r Plotting functions for shaded areas}
stat_sum_df <- function(fun, geom="crossbar", ...) {
  stat_summary(fun.data=fun, geom=geom, ...)
}

```

Initialize function to estimate number of ticks in plots
```{r Plotting functions for number of ticks}
number_ticks <- function(n) {function(limits) pretty(limits, n)}
```

Import shape file for maps
```{r Import shape file}
p <- shapefile("crete_gr.shp")
```

Import data
```{r Import data}
# read raster file of cell code values
b <- brick("cell_codes.asc")
s <- stack(b) # create stack starting with cell codes
# read ecological dataset (this is the richness for crete)
v <- read.csv(file="Crete_alien.csv", header=TRUE, sep=",")

# full absence-presence file for all spp
raw.all_spp <- read.table(file="crete_rawdata2.csv", sep="\t",
                          as.is = TRUE,  header=FALSE,
                          stringsAsFactors=FALSE)
raw.all_spp <- as.data.frame(t(raw.all_spp),
                             stringsAsFactors=FALSE)

# definition of alien spp
a <- read.table(file="alien_list.txt", as.is = TRUE, sep=",")

raw.alien_spp <- NULL
raw.alien_spp <- rbind(raw.alien_spp, raw.all_spp[,1])
# keep presenses only for alien spp
for (i in 2:dim(raw.all_spp)[2]){
  isalien <- which(a==raw.all_spp[1,i])
  if (length(isalien)>0){
    #raw.alien_spp[counter]<-raw.alien_spp[,-i]
    raw.alien_spp <- rbind(raw.alien_spp, raw.all_spp[,i])
  }
}

cells.no_of_aliens <- NULL
cells.withaliens <- NULL

for (i in 2:dim(raw.alien_spp)[2]){
  if (sum(as.numeric(raw.alien_spp[-1,i])) >0){
    cells.withaliens <- c(cells.withaliens, 
                        as.numeric(raw.alien_spp[1,i]))
    cells.no_of_aliens <- c(cells.no_of_aliens,
                          sum(as.numeric(raw.alien_spp[-1,i])))
  }
}
``` 

Turn richness dataset to stack of rasters
```{r Turn richness dataset to stack of rasters}
if (!file.exists("alien.RData")){
  temp <- b
  temp[temp>0]<-NA # a temp raster the size of the code layer
  alien <- s
  
  for (i in 1:length(v)) {
    names(temp) <- names(v[i])       # name layer with variable name
    alien <- stack(alien, temp) 
  }
  
  cells.notempty <- as.vector(alien$cell_codes)
  cells.notempty <- cells.notempty[!is.na(cells.notempty)]
  
  for (val in cells.notempty){ # for each cell code
    line <- which(v[,1]==val)      # find cell code in dataset
    if (length(line)>0){
      for (i in 2:dim(alien)[3]) {# if the code exists
          alien[[i]][b==val] <- v[line,i-1]   # replace code with value
      }
    }
  }
  save(alien, file = "alien.RData") #save for later
}else{
  load("alien.RData")
}
```
Turn presense/absence dataset to stack of rasters
```{r Turn presense/absence dataset to stack of rasters}
# fill stack with the values from the presence/absence dataset 
# taking addresses from the cell code layer
if (!file.exists("presence.RData")){
  presence <- s
  for (i in 2:dim(raw.alien_spp)[1]) {
    names(temp) = raw.alien_spp[i,1] # name layer with variable name
    presence <- stack(presence, temp) 
  }
  for (i in 2:dim(raw.alien_spp)[1]) {
    for (val in cells.withaliens){ # for each cell code
      line = which(as.numeric(raw.alien_spp[1,-1])==val)      # find cell code in dataset
      if (length(line)>0){          # if the code exists
        presence[[i]][b==val]=as.numeric(raw.alien_spp[i,line])   # replace code with value
      }
    }
  }
  save(presence, file = "presence.RData") #save for later
}else{
  load("presence.RData")
}
```
Turn individual spp presense into alien presence
```{r Turn individual spp presense into alien presence}
alien_10k <- stackApply(subset(presence,2:nlayers(presence)), 
                        nlayers(presence), fun=sum )
names(alien_10k) <- "alien_10k"

if (!any(names(alien)=="alien_10k")){
    alien <- stack(alien, alien_10k) # add to original stack
}
```
Estimate the aggregated predictor
```{r Estimate the aggregated predictor}
presence_20k <- disaggregate(aggregate(presence, fact = 2 , fun = any, na.rm = TRUE ), 
                  fact=2)
alien_20k <- stackApply(subset(presence_20k,2:nlayers(presence_20k)), 
                        nlayers(presence_20k), fun=sum )

#manage differences in extent of resampled raster
alien_20k<-crop(alien_20k, extent(alien_10k))
names(alien_20k) <- "alien_20k"


if (!any(names(alien)=="alien_20k")){
    alien <- stack(alien, alien_20k) # add to original stack
}

save(alien, file = "alien.RData") #save for later
```
Alien input-output plots
```{r Alien input-output plots, fig.width=12, fig.height=3.5}
load("alien.RData") # load the data in case you dont want to estimate it again

# do some plotting to make sure data has been read correctly
tiff("Figure_1.tiff", width = 12, height = 3.5, units = 'in', res = 300)
par(mfrow=c(1,2))
plot_map(p, alien$alien_10k, "Alien ssp presence 8.25 x 8.25 km")
plot_map(p, alien$alien_20k, "Alien spp presence 16.5 x 16.5 km")
dev.off()
```
Prepare data for bootstrap Random Forest
```{r Prepare data for bootstrap Random Forest}
# make a mask from altitude raster in order to clean up data from non-ground values
data_temp <- as.data.frame((alien), xy = TRUE)
ind.notempty <- which(with(data_temp, data_temp$Alt>=0))
data <- data_temp[ind.notempty,]

ind.keep <- c("alien_10k", "alien_20k", "Hab_Div", "Artif_Hab", "Agric_Hab",
              "Nat_Hab", "Endemic_Spp_Richness", "Natura", "Alt", "Alt_Range",
              "Soil_Div", "Temp_Mean", "Temp_Warm", "Temp_Cold", "Temp_Range", 
              "Precip_Mean", "Precip_Wet", "Precip_Dry", "Precip_Range", 
              "Artif_Cover", "Agric_Cover", "Natural_Cover")

data <- data[ind.keep]
data$alien_10k[data$alien_10k>0] <- 1
data$alien_20k[data$alien_20k>0] <- 1
data$alien_10k <- as.factor(data$alien_10k) 
data$alien_20k <- as.factor(data$alien_20k) 

save(data, file = "data.RData") #save for later
```
Run bootstrap Random Forest
```{r Run bootstrap Random Forest}

if (!file.exists("results.RData")){
  start.time <- Sys.time()
  
  load("data.RData") # load the data 
  # uncomment to have replicatable results
  #set.seed(12345)
  temp_frame <- numeric(4)
  
  temp_frame$id <- NULL
  temp_frame$cutoff <- NULL
  temp_frame$value <- NULL
  temp_frame$group <- NULL
  temp_partial <- NULL
  
  ind.iterations <- 100
  rf.importance2 <- NULL
  rf.importance7 <- NULL
  rf.gini2 <- NULL
  rf.gini7 <- NULL
  
  rf.partial2 <- NULL
  rf.partial2 <- NULL
  rf.partial7 <- NULL
  rf.partial7 <- NULL
  
  # bootstrap - this will take a while
  for (i in 1:ind.iterations){
    # break data in training and testing 
    ind.train = sample(2,nrow(data), replace=TRUE, prob=c(0.7, 0.3)) 
    data_train = data[ind.train==1,] #70% training 
    data_test = data[ind.train==2,]  #30% testing
    
    cutoff_intervals <- seq(0.1, 0.9, by = 0.1) # 0.05 does not work that well
    result.TN <- numeric(length(cutoff_intervals)) # correct rejection
    result.FN <- numeric(length(cutoff_intervals)) # miss (type ii)
    result.FP <- numeric(length(cutoff_intervals)) # false alarm (type i)
    result.TP <- numeric(length(cutoff_intervals)) # hit
    
    ind.counter <- 0
    
    for (ind.cutoff in cutoff_intervals){
      fit <- randomForest(alien_10k ~.,
                        data=data_train, 
                        importance=TRUE, 
                        do.trace=FALSE,
                        mtry=2,
                        ntree=10000, 
                        cutoff = c(1-ind.cutoff, ind.cutoff))
      
      ind.counter <- ind.counter + 1  
      
      prediction <- predict(fit, data_test, type = "response") # predict responce
      
      result_temp <- confusionMatrix(prediction,data_test$alien_10k, positive = '1')
      
      result.TN <- result_temp$table[1]
      result.FN <- result_temp$table[2]
      result.FP <- result_temp$table[3]
      result.TP <- result_temp$table[4]
    
      result.P <- result.TP+result.FN
      result.N <- result.TN+result.FP
     
      if (result.N >0 && result.P>0){
        temp_frame <- rbind(temp_frame, c(ind.counter, ind.cutoff,result.TN, "TN"))
        temp_frame <- rbind(temp_frame, c(ind.counter, ind.cutoff,result.FN, "FN"))
        temp_frame <- rbind(temp_frame, c(ind.counter, ind.cutoff,result.FP, "FP"))
        temp_frame <- rbind(temp_frame, c(ind.counter, ind.cutoff,result.TP, "TP"))
     
        if (ind.counter == 2){
        rf.importance2 <- cbind(rf.importance2, importance(fit)[,3] )
        rf.gini2 <- cbind(rf.gini2, importance(fit)[,4] )
          for(column in names(data_test)){ 
            pp <- partialPlot(fit, data_test, x.var=toString(column), plot = F)
            rf.partial2[[column]]$x <- c(rf.partial2[[column]]$x, t(pp$x))
            rf.partial2[[column]]$y <- c(rf.partial2[[column]]$y, t(pp$y))
          }
        }
        if (ind.counter == 7){
          rf.importance7 <- cbind(rf.importance7, importance(fit)[,3] )
          rf.gini7 <- cbind(rf.gini7, importance(fit)[,4] )
          for(column in names(data_test)){ 
            pp <- partialPlot(fit, data_test, x.var=toString(column), plot = F)
            rf.partial7[[column]]$x <- c(rf.partial7[[column]]$x, t(pp$x))
            rf.partial7[[column]]$y <- c(rf.partial7[[column]]$y, t(pp$y))
          }
        }      
      }
    }
  }
  
  results <- data.frame(temp_frame[-1,])
  names(results) <- c("id", "cutoff", "value", "group")
  
  results$group <- as.character(results$group)
  results$cutoff <- as.numeric(results$cutoff)
  results$value <- as.numeric(results$value)
  results$id <- as.numeric(results$id)
  
  save(results, file = "results.RData") #save for later
  save(rf.partial2, file = "rfpartial2.RData")
  save(rf.partial7, file = "rfpartial7.RData")
  
  save(rf.importance2, file = "rfimportance2.RData")
  save(rf.importance7, file = "rfimportance7.RData")
  
  save(rf.gini2, file = "rfgini2.RData")
  save(rf.gini7, file = "rfgini7.RData")
  
  end.time <- Sys.time()
  time.taken <- end.time - start.time
  time.taken
}else{
  load('results.RData')
  load('rfpartial2.RData')
  load('rfpartial7.RData')
  load('rfimportance2.RData')
  load('rfimportance7.RData')
  load('rfgini2.RData')
  load('rfgini7.RData')
}

```
Plot importance and genie index
```{r Plot importance and genie index}
formal.names <- c("alien_20k", 
                  "Habitat Diversity", "Artificial Habitat", "Agricultural Habitat",
                  "Natural Habitat", "Endemic spp richness", "Natura", 
                  "Altitude range", "Altitude", "Soil type diversity",
                  "Mean temperature", "Maximum temperature", "Minimum temperature", 
                  "Temperature range", "Mean precipitation", "Maximum precipitation",
                  "Minimum precipitation", "Precipitation range", "Artificial cover", 
                  "Agricultural cover", "Natural cover")

rf.importance7_use<-as.data.frame(t(rf.importance7))
colnames(rf.importance7_use)<-formal.names
mns <- colMedians(as.matrix(rf.importance7_use), na.rm=TRUE)
rf.importance7_use <- rf.importance7_use[,order(mns)]
rf.importance7_use<-rf.importance7_use[!names(rf.importance7_use)=="alien_20k"] #exclude alien_20k

rf.gini7_use<-as.data.frame(t(rf.gini7))
colnames(rf.gini7_use)<-formal.names
mns <- colMedians(as.matrix(rf.gini7_use), na.rm=TRUE)
rf.gini7_use <- rf.gini7_use[,order(mns)]
rf.gini7_use<-rf.gini7_use[!names(rf.gini7_use)=="alien_20k"] #exclude alien_20k

tiff("Figure_2.tiff", width = 12, height = 6, units = 'in', res = 300)
rf.gini7_long <- gather(rf.gini7_use, type, value, factor_key=TRUE)
rf.importance7_long <- gather(rf.importance7_use, type, value, factor_key=TRUE)

p_gimp.1 <- ggplot(rf.importance7_long, aes(type, value)) +
  theme_bw()+
  theme(axis.ticks.length = unit(-1 , "pt"))+
  theme(axis.text.x = element_text(margin=margin(5,5,5,5,"pt")))+
  theme(axis.text.y = element_text(margin=margin(5,5,5,5,"pt")))+
  geom_boxplot() +
  coord_flip()+
  ylab("Mean Decrease in Accuracy")+
  xlab("") 

p_gimp.2 <- ggplot(rf.gini7_long, aes(type, value)) + 
  theme_bw()+
  theme(axis.ticks.length = unit(-1 , "pt"))+
  theme(axis.text.x = element_text(margin=margin(5,5,5,5,"pt")))+
  theme(axis.text.y = element_text(margin=margin(5,5,5,5,"pt")))+
  geom_boxplot() +
  coord_flip()+
  ylab("Mean Decrease Gini")+
  xlab("") 

grid.arrange(p_gimp.1, p_gimp.2, ncol=2, nrow=1)
dev.off()
```
Plot partials comparison
```{r Plot partials comparison}
tiff("Figure_3.tiff", width = 12, height = 6, units = 'in', res = 300)
p_part.11 <- plot_partial(as.data.frame(rf.partial7$Natural_Cover), 542:590, 
                          "", "Natural Cover [%]")
p_part.12 <- plot_partial(as.data.frame(rf.partial7$Agric_Cover), 539:587, 
                          "", "Agricultural cover [%]")
p_part.13 <- plot_partial(as.data.frame(rf.partial7$Endemic_Spp_Richness), 531:553, 
                          "", "Endemic spp richness")
p_part.21 <- plot_partial(as.data.frame(rf.partial7$Alt_Range), 515:558, 
                          "Alien spp presence", "Altitude")
p_part.22 <- plot_partial(as.data.frame(rf.partial7$Alt), 502:545, 
                          "", "Altitude range [m]")
p_part.23 <- plot_partial(as.data.frame(rf.partial7$Temp_Mean), 555:581,
                          "", expression(paste("Mean temperature", ~degree*C)))
p_part.31 <- plot_partial(as.data.frame(rf.partial7$Precip_Mean), 549:580, 
                          "", "Mean precipitation [mm]")
p_part.32 <- plot_partial(as.data.frame(rf.partial7$Precip_Range), 545:577, 
                          "", "Precipitation range [mm]")
p_part.33 <- plot_partial(as.data.frame(rf.partial7$Natura), 546:585, 
                          "", "Natura cover [%]")

grid.arrange(p_part.11, p_part.12, p_part.13, 
             p_part.21, p_part.22, p_part.23,
             p_part.31, p_part.32, p_part.33,
             ncol=3, nrow=3)
dev.off()
```
Plot statistics vs cutoff
```{r Plot stats vs cutoff, fig.width=12, fig.height=6}
ind.counter <- 1

for(i in seq_len(nrow(results))){
  results$id[i] <- ind.counter
  if (i/4==floor(i/4)){
    ind.counter <- ind.counter + 1
  }
}

results.wide <- spread(results, group, value)
results.wide$TNR <- results.wide$TN / (results.wide$TN + results.wide$FP)
results.wide$TPR <- results.wide$TP / (results.wide$TP + results.wide$FN)

N = results.wide$TN + results.wide$TP + results.wide$FN + results.wide$FP
S = (results.wide$TP + results.wide$FN)/N
P = (results.wide$TP + results.wide$FP)/N
results.wide$MCC = (results.wide$TP/N - S*P)/sqrt(P*S*(1-S)*(1-P))

results.wide <- results.wide[, !(colnames(results.wide) %in% c("FN","FP","TN","TP"))]
results.long <- gather(results.wide,group, value, TNR:MCC)

tiff("Figure_4.tiff", width = 12, height = 6, units = 'in', res = 300)
p<-ggplot(data=as.data.frame(results.long), 
          aes(x=cutoff, y=value, group=group, colour=group)) +
          stat_sum_df("mean_cl_boot",geom="smooth")+
          scale_y_continuous (breaks=number_ticks(10))+
          ylab("Value")+
          xlab("Cutoff value")+
          theme_bw(base_size = 18)+
          theme(axis.ticks.length = unit(-5 , "pt"))+
          theme(axis.text.x = element_text(margin=margin(10,10,10,10,"pt")))+
          theme(axis.text.y = element_text(margin=margin(10,10,10,10,"pt")))+
          theme(legend.position = c(.07, .7))
p$labels$colour<-""
print(p)
dev.off()
```
